Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-10-22 paper2_JNLS/paper.tex
NM

A recent series of laboratory \cite{bolles2019} and numerical investigations \cite{viotti2014} have demonstrated the emergence of anomalous wave statistics from abrupt variations in bottom topography. Since topographical variations are strictly one-dimensional, these studies can be viewed as offering a bare minimum set of conditions capable of generating anomalous waves. In particular, the more complex mechanism of focusing by 2D bottom topography absent. The studies thus offer a paradigm system, with emergent anomalous statistical properties similar to that seen in more complex systems, but in a tractable context that is amenable to analysis.

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda and Qi \cite{majdaqi2019} analyzed more severely truncated systems that were shown to exhibit many of the same anomalous properties while enjoying a more tractable, lower-dimensional structure.

@@

We consider the surface displacement $\eta(x,t)$ of unidirectional, shallow-water waves in a reference frame moving with the characteristic wave speed, $\xi = x - ct$. Here, $c = \sqrt{g \depth}$ is leading-order approximation to the wave speed (i.e.~from linear theory), where $g$ is gravity and $\depth$ is the local depth.
The leading-order dynamics are corrected to first order in small amplitude by the Korteweg–de Vries equation (KdV), which in dimensional form is given by \cite{whitham2011linear}
\begin{equation}
\label{KdV}
\eta_t + \frac{3 c}{2 \depth} \eta \eta_{\xi} + \frac{c \depth^2}{6} \eta_{\xi \xi \xi} = 0
\end{equation}
% Given by Whitham on p. 461, Eq. 13.98
Motivated by the experiments, we consider waves that originate from a region of constant depth, encounter an abrupt depth change, and continue into another region of constant depth. Thus, depth will be piecewise constant
\begin{align}
\depth = 
\begin{cases}
\dup \quad \mbox{if } x<0 \\
\ddn \quad \mbox{if } x>0
\end{cases}
\end{align}
Most often, we will consider waves moving into shallower depth, so that $\dup > \ddn$.

In the experiments, the randomized incoming wave-field is generated with a peak forcing frequency of $\freqp = 2$ Hz, which gives rise to the characteristic wavelength of $\lam = c/\freqp = \sqrt{g \depth} / \freqp$. Note that both the characteristic wave speed $c = c_{\pm}$ and wavelength $\lam = \lamupdn$ each take different values upstream and downstream of the ADC. We remark that experimental measurements indicate that $\etastd$ is nearly the same on both values of the ADC. Hence, we will not distinguish between upstream and downstream values of $\etastd$.
%See Table \ref{paramtable} for a summary of important parameters.

@@

Meanwhile, the dimensionless depth $\drat = {\depth}/{\dup}$ {\em does} change value across the ADC since the depth $\depth$ changes. Recall that the reference frame of \eqref{dimlessKdV} moves with the local wave speed via the variable $\xi = x-ct$ from \eqref{KdV}. Thus, the ADC is met at some time $T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. Therefore, we can regard $\drat$ is a piece-wise-constant function of dimensionless time
\begin{equation}
\label{dratpw}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } {t}<0 \\
\dratdn = {\ddn}/{\dup} 	&\quad \mbox{for } {t}>0
\end{cases}
\end{equation}
See Table \ref{paramtable} for a summary of these dimensionless parameters and their values in experiments.

@@

Remarkably, the TKdV system \eqref{TKdV} retains nearly the same Hamiltonian structure described in Section \ref{HamiltonianSection}, with the only modification being the inclusion of the projection operator \cite{bajars2013weakly, majda2019}. The piecewise defined Hamiltonian for TKdV is given by
\begin{equation}
\label{TruncHamiltonian}
\HLupdn = C_2 \dratupdn^{1/2} \, \Htwo[\uL] - C_3 \dratupdn^{-3/2} \, \Hthree[\uL]
\end{equation}
where $\Hthree$ and $\Htwo$ are defined exactly as before \eqref{H3H2}, but now are simply applied to the projected variable $\uL = \Proj u$.
Then TKdV \eqref{TKdV} can be expressed as
\begin{align}
\partial_t {\uL} = \partial_x \Proj \, \vard{\HLupdn}{\uL}
\end{align}
where the truncated symplectic operator is $\SympL = \partial_x \Proj$.
% Note: Bajars 2013 has some important details on the Hamiltonian on the truncated system.

>>>>

A recent series of laboratory \cite{bolles2019} and numerical investigations \cite{viotti2014} have demonstrated the emergence of anomalous wave statistics from abrupt variations in bottom topography. Since topographical variations are strictly one-dimensional, these studies can be viewed as offering a bare minimum set of conditions capable of generating anomalous waves. In particular, the more complex mechanism of focusing by 2D bottom topography is  absent. The studies thus offer a paradigm system, with emergent anomalous statistical properties similar to that seen in more complex systems, but in a tractable context that is amenable to analysis.

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda and Qi analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure \cite{majdaqi2019}.

@@

We consider the surface displacement $\eta(x,t)$ of unidirectional, shallow-water waves in a reference frame moving with the characteristic wave speed, $\xi = x - ct$. Here, $c = \sqrt{g \depth}$ is leading-order approximation to the wave speed (i.e.~from linear theory), where $g$ is gravity and $\depth$ is the local depth.
The leading-order dynamics are corrected to first order in small amplitude by the Korteweg–de Vries equation (KdV), which in dimensional form is given by \cite{whitham2011linear}
\begin{equation}
\label{KdV}
\eta_t + \frac{3 c}{2 \depth} \eta \eta_{\xi} + \frac{c \depth^2}{6} \eta_{\xi \xi \xi} = 0
\end{equation}
% Given by Whitham on p. 461, Eq. 13.98
Motivated by the experiments, we consider waves that originate from a region of constant depth, encounter an abrupt depth change, and continue into another region of constant depth. Thus, depth will be piecewise constant
\begin{align}
\depth = 
\begin{cases}
\dup \quad \mbox{if } x<0 \\
\ddn \quad \mbox{if } x>0
\end{cases}
\end{align}
Most often, we will consider waves moving into shallower depth, so that $\dup > \ddn$. Throughout this paper, we will use the subscript `-' to represent upstream variables and `+' for downstream variables.

In the experiments, the randomized incoming wave-field is generated with a peak forcing frequency of $\freqp = 2$ Hz, which gives rise to the characteristic wavelength of $\lam = c/\freqp = \sqrt{g \depth} / \freqp$. Note that both the characteristic wave speed $c = c_{\pm}$ and wavelength $\lam = \lamupdn$ take different values upstream and downstream of the ADC. We remark that experimental measurements indicate that $\etastd$ is nearly the same on both sides of the ADC. Hence, we will not distinguish between upstream and downstream values of $\etastd$.
%See Table \ref{paramtable} for a summary of important parameters.

@@

Meanwhile, the dimensionless depth $\drat = {\depth}/{\dup}$ {\em does} change value across the ADC since the depth $\depth$ changes. Recall that the reference frame of \eqref{dimlessKdV} moves with the local wave speed via the variable $\xi = x-ct$ from \eqref{KdV}. Thus, the ADC is met at some time $T_{ADC}$, and for simplicity we set $T_{ADC} = 0$. Therefore, we can regard $\drat$ as a piece-wise-constant function of dimensionless time
\begin{equation}
\label{dratpw}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } {t}<0 \\
\dratdn = {\ddn}/{\dup} 	&\quad \mbox{for } {t}>0
\end{cases}
\end{equation}
See Table \ref{paramtable} for a summary of these dimensionless parameters and their values in experiments.

@@

Remarkably, the TKdV system \eqref{TKdV} retains the Hamiltonian structure described in Section \ref{HamiltonianSection}, with the only modification being the inclusion of the projection operator \cite{bajars2013weakly, majda2019}. The piecewise defined Hamiltonian for TKdV is given by
\begin{equation}
\label{TruncHamiltonian}
\HLupdn = C_2 \dratupdn^{1/2} \, \Htwo[\uL] - C_3 \dratupdn^{-3/2} \, \Hthree[\uL]
\end{equation}
where $\Hthree$ and $\Htwo$ are defined exactly as before \eqref{H3H2}, but now are simply applied to the projected variable $\uL = \Proj u$.
Then TKdV \eqref{TKdV} can be expressed as
\begin{align}
\partial_t {\uL} = \partial_x \Proj \, \vard{\HLupdn}{\uL}
\end{align}
where the truncated symplectic operator is $\SympL = \partial_x \Proj$.
% Note: Bajars 2013 has some important details on the Hamiltonian on the truncated system.
